Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-07-10 extended_catalog/paper/extended_catalog.tex
Typos

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that validate this tool
and calculate the LAT's detection threshold to spatially extended
sources.  We then use this tool to test all sources in the two
year source catalog for extension and report on the discovery
of XXXXXX\todo{What number should be put
here???} new extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope. It has been surveying the $\gamma$-ray
sky since June 2008.  Fermi is well suited as a survey
instrument because of its broad energy coverage (20 \mev
to $>300$ \gev), wide field of view ($\sim 2.4 \sr$) and large effective
area ($\sim 8000 \cm^2$ at $>1 \gev$).

@@

The LAT collaboration has previously reported on the discovered
a variety of spatially extended supernova remnants (SNRs)
IC443 and W51C, pulsar wind nebulae (PWNe) MSH 15-52 and Vela X
(\cite{ic443,w51c,msh1552,velax}), and galaxies the Large Magellanic
Cloud and the Small Magalenic Cloud (\cite{lmc,smc}.  There are other
source classes which could possibly be spatially resolved by the LAT
including galaxy clusters, dark matter, and possibly new physics.

@@

Being able to spatially resolve a LAT source is important for several
reasons. Most importantly, the spatial morphology can be used to
help identify the source at other wavelengths. Because of the large
point spread function (PSF), there are often several possible source
counterparts.  Finding a matching morphology can uniquely identify
the source.

Because of the largely varying PSF of the LAT, the spatial and spectral
model of a source do not nicely decouple. Therefore, spatially resolving
a LAT source is also important for obtaining an unbiased spectrum.
Furthermore, correctly modeling a source's extension is important for
improved the overall model of the sky and removing excess residuals.
These residuals around catalog sources can lead to finding spurious nearby
point sources. This was found to be the case with the one year catalog
in the region surrounding the Large Magellanic Cloud (\cite{first_cat}).

\section{Analysis Method}

Morphological studies of sources in the \gev energy range
using the LAT are challenging because of the significantly energy
dependent PSF and because of systematic errors in the galactic diffuse
emission.  The LAT's PSF is determined at lower energies by multiple
scattering in the tracker and the 68\% containment radius of the PSF is
approximately 5.1\deg at 100 \mev  when averaged over instrument acceptance
and including photons which convert in either the thick or thin layers of
the tracker. The PSF approaches a limit given by the granularity of the
tracker channels and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but are limited by statistics. Therefore,
a detailed analysis procedure mused be used to maximize our
sensitivity.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool
provides a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is computed is maximized by fitting
parameters of the model.  The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position
and extension can simultaneously fit by maximizing the likelihood.

This analysis is not very feasible using the official LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center\cite{fssc}} because \gtlike is only able to
fit the spectral parameters of the source. The extension fitting would
have to be done as an external loop outside of the code. Typically, the
run time for a binned source analysis using \gtlike is several hours
so this loop would be especially time consuming. What has typically
been done to study an extended sources is to fix the position of the
extended source and then develop a profile of the likelihood as a
function of the extension.  This approach is not optimal because the
position and extension of the source should be simultaneously maximized
to find the best fit parameters.  Furthermore since this approach is
very computationally intensive, no large scale Monte Carlo effort can
be launched to validate it.

An alternate approach was developed for fitting spatially extended
sources.  An alternate maximum likelihood fitting package called
\pointlike has been developed by the Fermi collaboration and we have
added to it the ability to analyze spatially extended sources. \pointlike
is described in the one year catalog and in Matthew Kerr's Ph.D thesis
(\cite{first_cat,matthew_kerr_thesis}).

What differentiates \pointlike most significantly from \gtlike is its
emphasis on speed by introducing approximations into the calculation
of the likelihood function.  This is done while at the same time
trying to minimized the introduced numerical errors.  Like the binned
version of \gtlike, \pointlike bins the sky both in position and energy.
Unlike \gtlike, \pointlike relies on the {\em healpix} representation of
spatial bins (\cite{healpix}). Since the PSF changes significantly with
energy, the spatial size of each bin for a given energy is calculated
to be smaller than than the PSF. This ensures that little information
is lost at each energy level while minimizing the overall number of bins.

>>>>

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies to validate this tool
and calculate the LAT's detection threshold to spatially extended
sources.  We then apply this tool to test all sources in the two
year source catalog for extension and report on the discovery
of XXXXXX\todo{What number should be put
here???} new extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope. It has been surveying the $\gamma$-ray
sky since June 2008.  Fermi is well suited as a survey
instrument because of its broad energy coverage (20 \mev
to $>300$ \gev), wide field of view ($\sim 2.4 \sr$), and large effective
area ($\sim 8000 \cm^2$ at $>1 \gev$).

@@

The LAT collaboration has previously reported on the discovered
a variety of spatially extended including the supernova remnants (SNRs)
IC443 and W51C (\cite{ic443,w51c}),
the pulsar wind nebulae (PWNe) MSH 15-52 and Vela X
(\cite{msh1552,velax}), and the Large Magellanic
and the Small Magalenic Cloud (\cite{lmc,smc}).  There are other
source classes which could possibly be spatially resolved by the LAT
including galaxy clusters, dark matter, and possibly new physics.

@@

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large point spread
function (PSF), there are often several possible source counterparts.
Finding a matching morphology can be used to uniquely identify the source.
Because of the largely varying PSF of the LAT, the spatial and spectral
information about a source do not nicely decouple. Therefore, a biased
spatial model will bias the spectral model of the source and studyign
the extension of the source can correct for htis.  Furthermore, correctly
modeling a source's extension is important for improving the overall model
of the sky and removing excess residuals.  These residuals manifested
themseleves in the one year catalog as spurious point sources in the
region region surrounding the Large Magellanic Cloud (\cite{first_cat}).

\section{Analysis Methods}

Morphological studies of sources in the \gev energy range
using the LAT are challenging because of the significantly energy
dependent PSF and because of systematic errors in the galactic diffuse
emission.  The LAT's PSF is determined at lower energies by multiple
scattering in the tracker and the 68\% containment radius of the PSF is
approximately 5.1\deg at 100 \mev  when averaged over instrument acceptance
and including photons which convert in either the thick or thin layers of
the tracker. The PSF approaches a limit given by the granularity of the
tracker channels and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but are typically limited by statistics. 
A detailed analysis procedure must be used to maximize our
sensitivity over all energies.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool
provides a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is maximized by fitting
parameters of the model.  The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position
and extension can be simultaneously fit.

This analysis is not feasible using the official LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center (\cite{fssc})} because \gtlike can only 
fit the spectral parameters of the model. The extension fitting would
have to be done as an external loop. Typically, the
run time for a binned source analysis using \gtlike is several hours
so this loop would be especially time consuming. What has typically
been done to study an extended sources is to fix the position of the
source and develop a profile of the likelihood as a
function of extension.  This approach is not optimal because the
position and extension of the source should be simultaneously maximized
to find the best fit parameters and to correctly compute the statistical
significance of the detection.  Furthermore since this approach is
computationally intensive, no large scale Monte Carlo effort can
be run to validate it.

We developed an alternate approach to fit spatially extended
sources. A separate maximum likelihood fitting package called
\pointlike has been developed by the LAT collaboration and we have
added to it the ability to analyze spatially extended sources. \pointlike
is described in the one year catalog paper and in Matthew Kerr's Ph.D thesis
(\cite{first_cat,matthew_kerr_thesis}).

What differentiates \pointlike most significantly from \gtlike is its
emphasis on speed by introducing approximations into the calculation
of the likelihood function while trying to minimized the introduced
numerical errors.  Like the binned version of \gtlike, \pointlike bins
the sky in both position and energy.  Unlike \gtlike, \pointlike relies on
the {\em healpix} representation of spatial bins (\cite{healpix}). Since
the PSF changes significantly with energy, the spatial size of each bin
is calculated to be smaller than than the PSF. By changing the spatial
bin size with energy, \pointlike ensures that little information is lost
at each energy while minimizing the required number of bins.
